\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-03-25 reviews/response-2.tex

These were not stated explicitly anywhere.
We've added them to table 1.
\item \textcolor{blue}{The boundary conditions are not discussed very rigorously systematically - the authors seem to
sprinkle in some boundary conditions here and there. I think the boundary conditions need to be
given for each of the models at the time the models are presented - boundary conditions are needed
to complete the definition of each models.}
\item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
there is no expression given for the strains $\dot\epsilon(u)$.
This is one of the things I cam across that need to
be made more precise.}

This was stated in table 1 but we've added the definition as $\frac{1}{2}(\nabla u + \nabla u^\top)$ and added a sentence to the text referring readers to table 1.
\item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
in the hybrid approach. What order is typically used?}

Added the following text: ``Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.''
\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen's flow law come into the hybrid model?
I was expecting to see it there, but maybe I'm missing something.}
\item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
different models in different regions and couple them (a la the ISCAL method). Is this possible, or
something that the authors are thinking to add to their model/code?}

Implementing this idea will require some new developments to Firedrake, namely first-class support for subdomains, defining different PDEs on different subdomains, and defining matching conditions for the solutions at the interfaces between subdomains.
The Firedrake developers are working on this feature right now as it's very much in demand.
We are very much interested in, for example, using SIA in the interior of the ice sheet and SSA in the ice streams and margins, as this would give a much less computationally-intensive way to do some form of whole-ice sheet modeling than using, say, the first order model.
\item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
Glen's law expression as well, since it is mentioned.}

We have added some text and equations describing the model we used, the boundary conditions, and the shear heating rate, the latter of which implicitly includes Glen's flow law.
\item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
is discretized. In typical ice sheet models, this equation is used to change the ice extent - one
meshes up a region of ``potential'' ice, and then uses the thickness to dynamically determine a mask
for ice-covered regions. Do you do something like this in your model? It should be discussed for
completeness. I think you maybe start to do this in Section 4.1, but it is very confusing and hard to
make the connection.}
\item \textcolor{blue}{What sort of meshes do you use in your model (in the horizontal dimension, for the hybrid one)?
Structured/unstructured? Hex/tet (quad/tri)?}

We use unstructured triangular meshes although Firedrake in principle can use unstructured quad meshes.
\item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
Is it to obtain parameters in the model like the basal friction using observational data of e.g. surface velocity? There really needs to be more discussion here, and I personally would like to see
a mathematical statement of a representative inverse problem you are solving. It would be worth
citing the work Perego et al. on optimization-based inversion, if what you are doing is similar:
https://doi.org/10.1002/2014JF003181. BTW, the basal friction has not been defined, yet it is
discussed - it needs to be defined earlier, when talking about boundary conditions (which needs to
be added).}

We added a brief description of the mathematics of the inverse problem to be solved.
We also added: ``The state to be estimated can be any single input field to the diagnostic model -- basal friction, rheology, or another field that the user has added by customizing the model.''

\item \textcolor{blue}{Section 4.1: I find this section confusing. I assume you are talking about discretizing the thickness
equation here - that should be made clear. I disagree with several statements in this section as
well. ``The simplest explicit timestepping schemes are unstable with CG finite elements'' - if you are
talking about CFL stability, this is not true. You need to satisfy a CFL condition which could give
rise to very small time-steps but you can get the scheme to be stable. I'm also confused about the
notion of SUPG as a time-stepping scheme - I think of SUPG as a finite element approach to deal
with advection-dominated flow problems, for example, that does not have anything to do with time-
stepping. Maybe you are referring to upwinding? In any case, I think SUPG has nothing to do with
forward Euler, so the discussion about forward Euler requiring parameters is erroneous. Additionally,
I don't understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
explicit and implicit Euler have effectively the same diffusion and dispersion properties, so there
should not really be a difference between the schemes. Did the authors verify their time-stepper on
a manufactured problem to ensure that it was implemented correctly?}
\item \textcolor{blue}{Section 4.2: there is an approach discussed in Tezaur et al. (https://doi.org/10.5194/gmd-8-1197-2015)
for dealing with bad initial conditions in a Newton solver that relies on homotopy continuation that
would be worth citing. It is an alternate to the approach you describe that lets you get away with not
doing a line search for Newton. By the way, it should be no surprise that Newton is not converging
without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
guess without the line search.}

@@

Added a reference to the Albany paper as well as another one on trust region methods.
We described a fairly rudimentary line search method in more detail than perhaps is necessary for a reader who's a seasoned modeler.
It's a bit of a pet peeve of mine when papers just say ``We used Newton!'' without any attention to the globalization strategy, which can make a huge difference to the solver robustness.
\item \textcolor{blue}{Section 4.3, lines 411-412: there are actually ways to construct weighted norms to deal with the issue
of DOFs having different orders of magnitude for the purpose of convergence.}

I think you can use a lumped mass matrix as the $H_0$ in BFGS too, but I find it to be far preferable to use something like Gauss-Newton which gives mesh-independent convergence and which achieves close to the second-order rate of full Newton on many problems.
No change to the text.
\item \textcolor{blue}{Section 4.4: Again, it is not clear to me what is your inverse problem. You need to state this explicitly
so it is clear.}

See correction to section 3.
The point of the inverse solver class is that it is very general with respect to what field is being inferred and what diagnostic model is being used.
Since the solver was designed to work for many different inverse problems
\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on ``extruded geometries'' - do you ever use non-
extruded geometries? It has been shown in various references that there can be numerical problems
for land-ice solvers that do not use extruded geometries, e.g. Tezaur et al.}

We have restricted our implementation to use only extruded geometries.
Having made this choice, we might not ever be able to solve really geometrically complex problems like what the Elmer/Ice crowd did with the drainage of the lake underneath T\^ete Rousse glacier.
But the simplifications that this results in for the vast majority of glaciological applications are so advantageous that it's a sacrifice we're willing to accept.
\item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that
discussion.}
\item \textcolor{blue}{Section 4.6: The authors mention running their code on 1 core. Is the code parallel - can it be run
on multiple cores? Are there any hope for performance portability of the code to take advantage of
emerging HPC architectures, e.g., GPUs?}

### 2021-03-25 icepack.tex
Change Marshall ref to Huybrechts 2002
More responses to reviewer 2

Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics \citep{joughin2014marine},
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood \citep{nick2010physically, werder2013modeling},
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{marshall2002north}.
\end{enumerate}
To accomplish these tasks, modeling tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
In this paper, we introduce a new Python software package for glacier flow modeling called \emph{icepack}.
Our goal is to make a tool that will be both accessible to novices and more productive for experts.

@@

\begin{table}[h]
    \begin{tabular}{l|l}
        Symbol & Meaning \\
        \hline
        $h$ & thickness \\
        $b$ & bed elevation  \\
        $s$ & surface elevation  \\
        $d$ & water depth \\
        $u$ & velocity \\
        $\dot a_s$ & surface mass balance \\
        $\dot a_b$ & basal mass balance \\
        $\dot\varepsilon$ & strain rate, $\frac{1}{2}(\nabla u + \nabla u^\top)$\\
        $C$ & bed friction coefficient \\
        $A$ & rheology coefficient \\
        $\nu$ & unit outward normal \\
        $J$ & action functional \\
        $\Omega$ & spatial domain \\
        $\Gamma$ & calving terminus
    \end{tabular}
    \caption{Mathematical symbols}
    \label{tab:mathematical-symbols}
\end{table}

\begin{table}[h]
    \begin{tabular}{l|l}
        Symbol & Meaning \\
        \hline
        $n$ & Glen's flow law exponent \\
        $m$ & Weertman sliding law exponent \\
        $\rho_I$ & ice density \\
        $\rho_W$ & seawater density \\
        $g$ & gravitational acceleration
    \end{tabular}
    \caption{Physical constants}
    \label{tab:physical-constants}
\end{table}

@@

We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
The enthalpy describes the heat content of the material in a way that incorporates both temperature and latent heat stored in meltwater.
The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
Using the enthalpy has the advantage of circumventing many of the difficulties associated with tracking the interface between cold and temperate ice.
Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.

The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
For one, while the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty.
It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
Second, other processes such as damage, fabric development, and impurities can affect the fluidity as well.
For these reasons, the diagnostic model takes in the fluidity as an argument and users must calculate the fluidity field themselves, either using the default parameterization that icepack supplies or one of their own.
Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.

@@

Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
\begin{itemize}
    \item the model object and the method that solves the diagnostic equation,
    \item the objective and regularization functionals,
    \item the observed field and the name of the argument to the diagnostic solver,
    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver, and
    \item extra data passed to the diagnostic solver such as boundary conditions.
\end{itemize}
This class currently assumes that the observed state is always the ice velocity.
In principle the same design would suffice for more complicated inverse problems and this is an area of active development.

@@

Existing work on glaciological inverse problems uses the mismatch between the computed and observed velocity fields as part of the objective functional.
The effect of thickness errors is studied largely through a posteriori validation \citep{joughin2004basal, larour2005rheology}.
Errors in thickness or surface slope can be large enough that it might be impossible to fit the velocity measurements to the degree that statistical theory predicts \citep{macayeal1995basal}.
The velocity measurements themselves might have significant outliers, in which case using the usual weighted sum of squared misfits as an error metric will give poor results.
For these reasons, some studies have explored alternative objective functionals \citep{morlighem2010spatial}.
We have opted to use the regularized $L^1$-type error metric
\begin{equation}
    E(u - u^o) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)dx.
    \label{eq:l1-error-metric}
\end{equation}
This error metric approaches the usual weighted sum of squared errors as $\gamma \to \infty$, and approaches the sparsity-promoting $L^1$ error metric as $\gamma \to 0$.
For finite, positive values of $\gamma$ this error metric is robust to non-normality or a small fraction of outliers \citep{barron2019general}.

>>>>

Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics \citep{joughin2014marine},
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood \citep{nick2010physically, werder2013modeling},
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{huybrechts2002sea}.
\end{enumerate}
To accomplish these tasks, modeling tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
In this paper, we introduce a new Python software package for glacier flow modeling called \emph{icepack}.
Our goal is to make a tool that will be both accessible to novices and more productive for experts.

@@

\begin{table}[h]
    \begin{tabular}{l|l}
        Symbol & Meaning \\
        \hline
        $h$ & thickness \\
        $b$ & bed elevation  \\
        $s$ & surface elevation  \\
        $d$ & water depth \\
        $u$ & velocity \\
        $\dot a_s$ & surface mass balance \\
        $\dot a_b$ & basal mass balance \\
        $\dot\varepsilon$ & strain rate, $\frac{1}{2}(\nabla u + \nabla u^\top)$\\
        $C$ & bed friction coefficient \\
        $A$ & rheology coefficient \\
        $E$ & enthalpy density \\
        $\nu$ & unit outward normal \\
        $J$ & action functional \\
        $\Omega$ & spatial domain \\
        $\Gamma$ & calving terminus
    \end{tabular}
    \caption{Mathematical symbols}
    \label{tab:mathematical-symbols}
\end{table}

\begin{table}[h]
    \begin{tabular}{l|l}
        Symbol & Meaning \\
        \hline
        $n$ & Glen's flow law exponent \\
        $m$ & Weertman sliding law exponent \\
        $\rho_I$ & ice density \\
        $\rho_W$ & seawater density \\
        $c$ & specific heat capacity of ice \\
        $L$ & latent heat of melting of ice \\
        $g$ & gravitational acceleration
    \end{tabular}
    \caption{Physical constants}
    \label{tab:physical-constants}
\end{table}

@@

We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
The enthalpy density $E$ describes the heat content of the material in a way that incorporates both temperature $T$ and latent heat stored in meltwater:
\begin{equation}
    E = \rho_I(cT + Lf)
\end{equation}
where $L$ is the latent heat of melting of ice and $f$ is the melt fraction.
The melt fraction can only be positive when the ice as at the pressure-melting point.
The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
Using the enthalpy has the advantage of circumventing many of the difficulties associated with tracking the interface between cold and temperate ice.
Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.
The resulting PDE is
\begin{equation}
    \left(\frac{\partial}{\partial t} + \nabla\cdot u\right)E - \frac{\partial}{\partial z}\alpha\frac{\partial E}{\partial z} = q
\end{equation}
where $u$ is the full 3D velocity, $\alpha$ the (temperature-dependent) thermal diffusivity, and $q$ the volumetric heat sources.
This form assumes some amount of vertical diffusion of water at the melting point.
A realistic treatment would require treating the ice as a porous medium and modeling englacial water transport explicitly, which we have not done.
The boundary condition at the ice base is a fixed flux of heat from geothermal sources or from the oceans.
We depart from \citet{aschwandend2012enthalpy} in one respect.
Rather than fix the ice surface temperature to the atmospheric temperature, we instead use a Robin boundary condition that makes the ice temperature adjust to the external temperature:
\begin{equation}
    -\alpha\frac{\partial E}{\partial z}\Big|_{z = s} = \kappa(E - E_s)
\end{equation}
where $\kappa$ is a surface exchange coefficient and $E_s$ is the temperature-equivalent enthalpy that forces the surface.
Using a Robin instead of a Dirichlet boundary condition at the surface allows for more gradual adjustment to above-freezing temperatures that might otherwise create areas of unrealistically high melt fraction.
Except in blue ice areas, the thermal contact between ice and the atmosphere is mediated by firn, which we do not model explicitly.
The Robin boundary condition is a compromise to work around this missing component of the system.
Finally, if there is an upstream boundary the user must provide the inflow values of the enthalpy with depth, while outflow boundary conditions are applied at the glacier terminus.

The heat transport model assumes that the user will supply a volumetric heating rate, but the model itself does not calculate the heating rate from other fields.
The simplest description for the shear heating rate is
\begin{equation}
    q = \tau : \dot\varepsilon = A^{\frac{-1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}.
\end{equation}
The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
While the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty \citep{cuffey2010physics}.
It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
Other processes such as damage, fabric development, and impurities can affect the fluidity as well.
Moreover, there may be other volumetric heat sources that users wish to account for, such as cryo-hydrologic warming \citep{phillips2010cryo}.
These considerations defy any attempt to have one function that calculates the volumetric heating rate from the other state variables.
Instead, we opted for an interface where users calculate this rate themselves, with tutorials and examples that show how to do so for simple scenarios, leaving the freedom to alter the thermal feedbacks as they see fit.
Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.

@@

Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
Mathematically, this inverse problem amounts to finding a critical point of the functional
\begin{equation}
    L(u, \theta, \lambda) = J(u) + R(\theta) + \langle F(u, \theta), \lambda\rangle
\end{equation}
where $J$ is the misfit between the computed velocity $u$ and observations, $\theta$ is the field to be inferred and $R$ a regularization functional that measures the spatial variability of this field, $F$ is the diagnostic physics, and $\lambda$ a Lagrange multiplier to enforce the physics \citep{macayeal1992basal, joughin2004basal, larour2005rheology, shapero2016basal}.
The optimal value $\theta$ gives the best fit to observations subject to the constraints of the physics and that it should not overfit to noise in the data.
The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
\begin{itemize}
    \item the model object and the method that solves the diagnostic equation,
    \item the objective and regularization functionals,
    \item the observed field and the name of the argument to the diagnostic solver,
    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver, and
    \item extra data passed to the diagnostic solver such as boundary conditions.
\end{itemize}
This class currently assumes that the observed state is always the ice velocity.
The state to be estimated can be any single input field to the diagnostic model -- basal friction, rheology, or another field that the user has added by customizing the model.
In principle the same design would suffice for more complicated inverse problems and this is an area of active development.

@@

Existing work on glaciological inverse problems uses the mismatch between the computed and observed velocity fields as part of the objective functional.
The effect of thickness errors is studied largely through a posteriori validation \citep{joughin2004basal, larour2005rheology}.
Errors in thickness or surface slope can be large enough that it might be impossible to fit the velocity measurements to the degree that statistical theory predicts \citep{macayeal1995basal}.
The velocity measurements themselves might have significant outliers, in which case using the usual weighted sum of squared misfits as an error metric will give poor results.
For these reasons, some studies have explored alternative objective functionals \citep{morlighem2010spatial}.
We have opted to use the regularized $L^1$-type error metric
\begin{equation}
    J(u) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)dx.
    \label{eq:l1-error-metric}
\end{equation}
This error metric approaches the usual weighted sum of squared errors as $\gamma \to \infty$, and approaches the sparsity-promoting $L^1$ error metric as $\gamma \to 0$.
For finite, positive values of $\gamma$ this error metric is robust to non-normality or a small fraction of outliers \citep{barron2019general}.
